Draft -- Discriminating Seagrasses From Green Macroalgae in European Intertidal areas using high resolution multispectral drone imagery -- Draft

### 2024-05-02 docs/_tex/index.tex
adding bede's comments

Before each flight, targets used as ground control points were
distributed over the study site and georeferenced with a Trimble © Geo
XH 6000 differential GPS (dGPS). Ground control points were used to
correct georeferencing imprecision of orthomosaics with an horizontal
and vertical accuracy of 10cm. A dGPS was also used to georeference
quadrats of 0.25 m² ,which assessed the presence or absence of five key
taxonomic classes of intertidal vegetation : Bacillariophyceae
(Unicellular benthic diatoms forming biofilms at the sediment surface
during low tide), Phaeophyceae (brown macroalgae), Magnoliopsida (dwarf
seagrass), Chlorophyceae (green macroalgae) and Rhodophyceae (red
macroalgae) (Figure~\ref{fig-vegetation}). Pictures of each quadrat were
uploaded online to the Global Biodiversity Information Facility (GBIF)
platform \citep{BedeGbif}, a public open portal to store and share
biodiversity data. Each photograph was also processed to estimate the
percent cover of each type of vegetation using an image processing
software \citep[ImageJ]{schneider2012nih}. Hyperspectral reflectance
signatures of each vegetation class were recorded using an ASD FieldSpec
HandHeld 2 spectroradiometer, which acquires reflectance between 325 and
1075 nm, with 1 nm of spectral resolution. Hyperspectral signatures
served dual purposes: they validate the radiometric calibration of drone
data and contribute to error reduction in photo interpretation.

@@

A structure-from-motion photogrammetry software (Agisoft Metashape) was
used to process images to obtain multispectral orthomosaics of each
flight. The process for orthomosaicking was identical for every flight.
First, tying key points were detected inside of each image and between
overlapping images in order to obtain a sparse point cloud. This cloud
was cleaned using reprojection accuracy metric in order to remove noisy
points. A dense point cloud was then produced using a structure from
motion algorithm. A surface interpolation of this dense point cloud was
made to obtain a digital surface model (DSM), used to reconstruct the
multispectral ortho-image \citep{nebel2020review}. Low altitude drone
flights produced ortho-images with a very high spatial resolution (8 mm
per pixel), making it efficient to visually distinguish between the
various types of vegetation. High altitude fights on the other hand
allowed to cover large areas and produced images with a pixel size of 80
mm (Table~\ref{tbl-flights}).

@@

@@

The model was applied to all flights at both 12 and 120 m of altitude.
In situ information on georeferenced class type and percent cover
collected during each flight was used to assess the model accuracy.
These images were used to construct a validation dataset indicating the
presence or absence of each class. Additionally to the quadrat-based
validation dataset, polygons of each class were photo interpreted in
order to increase the number of pixels of the validation dataset. A
confusion matrix, along with precision metrics such as global accuracy,
sensitivity, specificity, and Kappa coefficient, was generated for each
sites. All validation matrices were then aggregated to create a unique
matrix . Altogether, a total of 536,000 pixels was used to validate the
model, thus providing a geographically robust validation dataset.

\subsection{Variable Importance Plots
(VIP)}\label{variable-importance-plots-vip}

Variable Importance Plots (VIP) serve as a method to identify which
predictors are important for predicting a specific class. Out of the
twenty one predictors utilized in this study, Variable Importance in
Projection (VIP) was computed only for the raw and standardized values
of the 10 spectral bands captured by the MicaSense camera. This is
achieved by repeatedly predicting the same dataset while shuffling one
predictor at a time. The benchmark score obtained after each iteration
is then compared to the benchmark score obtained without shuffling any
variables. The greater the difference between these two benchmark
values, the more important the variable is for the model
\citep{WEI2015399}.

@@

Each drone flight was used to produce a prediction map, as well as a
probability map that indicates the likelihood of the selected class for
every pixel. The low-altitude flight conducted in Gafanha, Portugal,
represents the site with the highest complexity
(Figure~\ref{fig-GafLow}). Among the five vegetation classes on which
the model was trained, four were present on this site, with
Chlorophyceae and Rhodophyceae mixed with the seagrass meadow. There was
also Bacillariophyceae forming biofilms on parts of sediment surface.
Although the seagrass bed was solely composed of \emph{Nanozostera
noltei}, various colors of this specie could be observed from dark green
(corresponding to healthy leaves) to whitish/brown (when leaves were
discolored having lost their pigmentation). Regardless of color,
Magnoliopsida pixels were accurately predicted by the model.

@@

\caption{\label{fig-GafHigh}RGB orthomosaic (Left) and Prediction
(Right) of the high altitude flight of Gafanha, Portugal. The total
extent of this flight was about 1 km² with a resolution of 80 mm per
pixel. Background colors indicate intertidal area (Light Green), land
area (Light Grey) and water (Light Blue). The yellow polygon shows the
extent of the low altitude flight of Gafanha presented in
Figure~\ref{fig-GafLow}. The zoom covers an area equivalent to a
10-meter Sentinel-2 pixel size.}

@@

Amoung the high altitude flights, the one acquired over the inner lagoon
of the Ria de Aveiro covered the largest area with approximately 1.5 km²
(Figure~\ref{fig-Boat}). This site was dominated by seagrass and the red
macroalgae \emph{Gracilaria vermicullophyla}. The classification
provided consistent results, with a patchy seagrass meadow mixed with
red macroalgae on the eastern part of the site. As shown in the zoom
(Figure~\ref{fig-Boat}), the edges of the meadow were colonised by green
macroalgae

@@

The flight over L'Epine in Noirmoutier Island, France
(Figure~\ref{fig-Dike}) was conducted near a dike, which crossed the
northern part of the site from west to east. Alongside this dike, brown
algae for the order Fucales (\emph{Fucus spp.}, \emph{Ascophyllum
nodosum}) were attached to sparse rocks and stranded green algae
(\emph{Ulva spp.}) could be observed Despite the high mixture between
Chlorophyceae and Magnoliopsida these two classes were correctly
discriminated by the classifier.

@@

\caption{\label{fig-Dike}RGB orthomosaic (Top) and Prediction (Bottom)
of the Northern part of Noirmoutier Island, France. The total extent of
this flight was about 28 000 m² with a resolution of 80 mm per pixel.
Background colors indicate intertidal area (Light Green) and land area
(Light Grey). The zoom covers an area equivalent to a 10-meter
Sentinel-2 pixel size.}

@@

In this study, a key innovation lies in the utilization of drone flights
at two different altitudes (12m and 120m) for constructing the neural
network model. The lower altitude flights (8mm spatial resolution)
enabled precise selection of pure pixels to train the neural network
model. This methodology implies a consistency between the reflectance
measured from both altitudes. Figure~\ref{fig-CompareRef} depicts the
relationship between reflectance from a low and a high-altitude flights
conducted over the same area.

@@

Overall there was a good agreement between the two altitudes. There was
however a slight underestimation for raw reflectance values in the
high-altitude flight, particularly for higher reflectance values
(Figure~\ref{fig-CompareRef} A). Since both flights were conducted over
vegetation areas, the highest reflectance values correspond to the
infrared part of the spectrum. This different is not present anymore
when reflectance values of both flights have been standardized
(Equation~\ref{eq-std} ; Figure~\ref{fig-CompareRef} B).

@@

A total of 536,000 pixels were used to validate the Neural Network
classifier. The sites with the lowest and highest number of validation
data were Gafanha Low (17316 pixels) and Marinha Lanzarote (159713
pixels), respectively. Model global accuracy was 94.26\% with a Kappa
coefficient of 0.92 (Figure~\ref{fig-Validation}).

@@

\caption{\label{fig-Validation}A global confusion matrix on the left is
derived from validation data across each flight, while a mosaic of
confusion matrices from individual flights is presented on the right.
The labels inside the matrices indicate the balanced accuracy for each
class. The labels at the bottom of the matrices indicate the User's
accuracy for each class, and those on the right indicate the Producer's
Accuracy. The values adjacent to the names of each site represent the
proportion of total pixels from that site contributing to the overall
matrix. Grey lines within the mosaic indicate the absence of validation
data for the class at that site. The table at the bottom summarizes the
Sensitivity, Specificity, and Accuracy for each class and for the
overall model.}

@@

The lowest performing site was Gafanha High (global accuracy of 75.45\%)
whereas Mataducos was the site with the most accurate prediction (global
accuracy of 98.05\%). Overall, the classes Phaeophyceae, Magnoliopsida,
Sediment and Rhodophyceae were correctly classified with a balanced
accuracy of 1, 0.96, 0.96 and 0.91 respectively. Bacillariophyceae was
the least performing class (accuracy of 0.72 ) mainly due to the
confusion with Magnoliopsida and Sediment.

\subsection{Variable importance}\label{variable-importance}

@@

\caption{\label{fig-VIP}Variable Importance of the Neural Network
Classifier for each taxonomic class. The bigger the slice, the more
important the variable for prediction of each class. The right plot
shows the drone raw and standardised reflectance spectra of each class.
Each slice represents the VIP of both raw and standardised reflectance
combined.}

@@

The spectral bands at 444, 717 and 842 nm of the Micasense camera did
not provide as important information to discriminate any of the
vegetation classes. The band at 531 nm was the only important predictor
for the classifier to accurately predict Chlorophyceae. In fact, at this
wavelength, the Chlorophyceae spectra showed the highest reflectance
among of all vegetation classes. The bands at 531 and 740 nm were the
most important predictors for Phaeophyceae, corresponding to the lowest
reflectance among all classes. Bands at 475 and 560 nm were the most
important predictors for Bacillariophyceae and Rhodophyceae,
respectively. Four predictors, ranging from the green (560 nm) to the
RedEdge (705 nm) bands were important to accurately predict
Magnoliopsida.

@@

\caption{\label{fig-upscaling}Kernel density plot showing the proportion
of pixel well classified based on the percent cover of the class in high
altitude flight pixels of Gafanha, Portugal. Each subplot shows all the
pixels of the same classes on the high altitude flight. Percent cover of
classes was retrieved using the result of the classification of the low
altitude flight of Gafanha, Portugal. The vertical dashed line shows the
0.85 probability of the model. Everything on the right of this line has
a probability higher then 0.85 and everything on the left of this line
has a lower probability.}

@@

When the percent cover of a given class was 100 \%, coarser high-flight
pixels are well classified for all the classes except for Bare Sediment,
which was only well classified 80\% of the time. This phenomenon is
likely attributed to the time gap between the two flights, allowing for
microphytobenthos migration to surface during low tide, consequently
altering the model's classification from bare sediment to
Bacillariophyceae. A percent cover of at least 80\% was sufficient to
have all the pixels of high altitude flights correctly classified, with
the exception of Magnoliopsida that required a higher percent cover
(\textgreater90 \%) to be well classified. Concerning the probability of
each class, the highest percent cover was needed to confidently predict
Bacillariophyceae. To predict Chlorophyceae with a confidence of 0.85, a
percent cover of 93 \% was needed, 90 \% for Magnoliopsida, 92 \% for
Rhodophyceae and 97 \% for Bacillariophyceae.

@@

The primary objective of this study was to develop a method for the
accurate classification of macrophytes on intertidal mudflats,
specifically focusing on distinguishing between Chlorophyceae (green
macroalgae) and marine Magnoliopsida (seagrasses) using multispectral
drone data. The ability to differentiate between various types of
vegetation plays a critical role in ecological monitoring and coastal
management. By distinguishing between seagrasses and macroalgae, our
approach facilitates targeted conservation strategies, enabling more
effective preservation and restoration efforts in coastal ecosystems.
The discrimination of seagrasses from green macroalgae presents two main
challenges \citetext{\citealp[ ]{oiry2021using}; \citealp[
]{bannari2022}; \citealp{veettil2020opportunities}}. First these two
macrophytes have a similar pigment composition: chlorophyll-a (common to
all vegetation types), chlorophyll-b (an additional photosynthetic
pigment), and accessory carotenoids such as zeaxanthin, lutein and
neoxanthin (Figure~\ref{fig-Pigm}). Second, seagrass and green
macroalgae frequently co-occur in intertidal areas, and can intermingle
within a remote sensing pixel if the spatial resolution is too low.
Here, the issue of intra-pixel mixing was resolved thanks to the very
high spatial resolution of the drone, and the risk of spetral confusion
was avoided thanks to the efficiency of the multispectral classifier.
Our results confirm a recent study based on in situ radiometry, where it
was demonstrated that a radiometer with at least eight spectral bands
ranging from 500 to 850 nm and including a green band at 530 nm and a
RedEdge band at 730 nm was crucial to accurately discriminate green
macrophytes \citep{Davies2023}.

@@

Meeting these two criteria, the Micasense RedEdge-MX DUAL camera used in
this study, enabled the classifier to achieve 97\% of accuracy between
these two classes (Figure~\ref{fig-ValidationGreen}). Even if the
pigment composition of green macrophytes is similar, differences in the
spectral shape can be observed, with green algae having a higher
reflectance peak at 560 nm as well as a higher NIR plateau than seagrass
(Figure~\ref{fig-vegetation}). Such differences were previously
attributed to differences in pigments concentration and/or ratios
\citep{bargain2013seasonal}, cellular structure as well as in the
disposition of the plant over the sediment surface \citetext{\citealp[
]{beach1997vivo}; \citealp[
]{kirk1994light}; \citealp{hedley2018influence}}.

The VIP analysis (Figure~\ref{fig-VIP}) identified that the band at 531
nm was the most important for accurately identifying Chlorophyceae. In
fact, at this wavelength, Chlorophyceae exhibited the highest
reflectance among all other classes, highlighting the difference in
carotenoids to chlorophyll-a ratios between seagrasses and green
macroalgae \citep{repolho2017seagrass}.

Concerning Phaeophyceae, the thick cell walls of these macroalgae
\citep{charrier2021growth} make it more reflective in the infrared part
of the spectra \citep{Slaton2001} whereas the presence of fucoxanthin
and zeaxanthin result in a low reflectance in the visible region
(Figure~\ref{fig-Pigm} ; Figure~\ref{fig-VIP}). These two key features
have been identified by the Neural Network as the two principal
predictors to accurately identify brown algae (Figure~\ref{fig-VIP}).
Similarly, the presence of phycoerythrin and phycocyanin in Rhodophyceae
contributes to the lowest reflectance among all classes in the spectral
range from 560 to 615 nm (Figure~\ref{fig-VIP}). Indeed the band at 560
nm has been identified as important for identifying this class likely
due to phycoerythrin absorption at this wavelength. Regarding
Bacillariophyceae, the VIP analysis (Figure~\ref{fig-VIP}) indicated
that 475 nm was the most important predictor for this class. Indeed, the
reflectance at 475 nm was higher for Bacillariophyceae than for any
other vegetation class (Figure~\ref{fig-vegetation}), very likely due to
the low biomass (and associated concentration of blue-absorbing
pigments) of microphytobenthos biofilms compared to seagrass and
macroalgae.

@@

While comparing the reflectance of both altitudes (12 m and 120 m), it
was observed Figure~\ref{fig-CompareRef} that there is an
underestimation of the infrared part of the spectra in the high-altitude
dataset. Such disparity in infrared reflectance may stem from temporal
differences between the flights, possibly resulting in a slightly drier
intertidal area and consequently higher infrared reflectance. This
disparity poses an issue for the methodology followed in the present
study, relying solely on one flight height for training. To address this
issue, we employed min/max standardized reflectance spectra as
predictors for the model (Equation~\ref{eq-std} ; \citep{Cao2017}). This
approach allowed us to eliminate the slight reflectance difference
between the flights (Figure~\ref{fig-CompareRef} B) and to focus on the
shape of the spectra in the visible part of the electromagnetic spectra,
where different pigmentation are associated to taxonomic dignostic
features. This was a key feature in building a model that could reliably
predict vegetation across geographical sites and seasons. It enabled
consistent prediction of vegetation classes across variations in biomass
and variability in light conditions \citetext{\citealp[
]{fyfe2003spatial}; \citealp[
]{COSTA2021107018}; \citealp{piaser2023impact}}.

The Figure~\ref{fig-upscaling} demonstrates that at least 90 \% of
seagrass cover is necessary for confident prediction of its presence.
This highlights a limitation of the methodology used to construct the
training dataset for the model. The dataset used to train our algorithm
was composed exclusively of pure pixels, which has resulted in the
model's reduced confidence when faced with lower percentages of seagrass
cover. Also, intertidal seagrasses exhibit marked phenology, with
varying pigment composition throughout the year\citetext{\citealp[
]{bargain2013seasonal}; \citealp{legare2022remote}}. Since the training
dataset has been made using well developed seagrass meadows, this model
may be less accurate outside of the seasonal seagrass peak of biomass.
Further investigation is required to evaluate the accuracy of the method
along different periods of the year.

@@

Climate change, global warming, eutrophication, alien and invasive
species development, coastal erosion, and sealevel rise are expected to
continue impacting coastal ecosystems in the future
\citetext{\citealp{SCHIBALSKI2022101414}; \citealp[
]{holon2018predictive}; \citealp{marquet2024global}}. and the demand for
meaningful and efficient monitoring of coastal habitats has never been
higher\citetext{\citealp[ ]{muller2018satellite}; \citealp[
]{villalobos2023remote}; \citealp{oiry2021using}}. Our findings,
particularly the improved discrimination of seagrass and green
macroalgae from other intertidal vegetation classes, highlight the
potential of drone-based remote sensing to support diverse applications,
from biodiversity conservation to climate change adaptation strategies.

>>>>

Before each flight, targets used as ground control points were
distributed over the study site and georeferenced with a Trimble © Geo
XH 6000 differential GPS (dGPS). Ground control points were used to
correct georeferencing imprecision of orthomosaics with an horizontal
and vertical accuracy of 10cm. A dGPS was also used to georeference
quadrats of 0.25 m² ,which assessed the presence or absence of five key
taxonomic classes of intertidal vegetation : Bacillariophyceae
(Unicellular benthic diatoms forming biofilms at the sediment surface
during low tide), Phaeophyceae (brown macroalgae), Magnoliopsida (dwarf
seagrass), Chlorophyceae (green macroalgae) and Rhodophyceae (red
macroalgae) (Figure~\ref{fig-vegetation}). Pictures of each quadrat were
uploaded online to the Global Biodiversity Information Facility (GBIF)
platform \citep{BedeGbif}, a public open portal to store and share
biodiversity data. Each photograph was also processed to estimate the
percent cover of each type of vegetation using an image processing
software \citep[ImageJ,][]{schneider2012nih}. Hyperspectral reflectance
signatures of each vegetation class were recorded using an ASD FieldSpec
HandHeld 2 spectroradiometer, which acquires reflectance between 325 and
1075 nm, with 1 nm of spectral resolution. Hyperspectral signatures
served dual purposes: they validate the radiometric calibration of drone
data and contribute to error reduction in photo interpretation.

@@

A structure-from-motion photogrammetry software \citep[Agisoft
Metashape,][]{agisoft}was used to process images to obtain multispectral
orthomosaics of each flight. The process for orthomosaicking was
identical for every flight. First, tying key points were detected inside
of each image and between overlapping images in order to obtain a sparse
point cloud. This cloud was cleaned using reprojection accuracy metric
in order to remove noisy points. A dense point cloud was then produced
using a structure from motion algorithm. A surface interpolation of this
dense point cloud was made to obtain a digital surface model (DSM), used
to reconstruct the multispectral ortho-image \citep{nebel2020review}.
Low altitude drone flights produced ortho-images with a very high
spatial resolution (8 mm per pixel), making it efficient to visually
distinguish between the various types of vegetation. High altitude
fights on the other hand allowed to cover large areas and produced
images with a pixel size of 80 mm (Table~\ref{tbl-flights}).

@@

\subsubsection{Model building}\label{model-building}

A neural network classification model was built using the fastai
workflow \citep{howard2018fastai}. This model is composed by 2 hidden
layers and have a total of 26 054 trainable parameters. Parameters have
been find tune using 12 epoch to minimize the error rate.

@@

The model was applied to all flights at both 12 and 120 m of altitude.
In situ information on georeferenced class type and percent cover
collected during each flight was used to assess the model accuracy.
These images were used to construct a validation dataset indicating the
presence or absence of each class. Additionally to the quadrat-based
validation dataset, polygons of each class were photo interpreted in
order to increase the number of pixels of the validation dataset.A total
of 536,000 pixels were used to validate the Neural Network classifier.
The sites with the lowest and highest number of validation data were
Gafanha Low (17316 pixels) and Marinha Lanzarote (159713 pixels),
respectively.A confusion matrix, along with precision metrics such as
global accuracy, sensitivity, specificity, and Kappa coefficient, was
generated for each sites. All validation matrices were then aggregated
to create an overall matrix .

\subsection{Variable Importance}\label{variable-importance}

Variable Importance Plots (VIP) serve as a method to identify which
predictors are important for predicting a specific class. Out of the
twenty one predictors utilized in this study, Variable Importance was
computed only for the raw and standardized values of the 10 spectral
bands captured by the MicaSense camera. This is achieved by repeatedly
predicting the same dataset while randomly shuffling one predictor at a
time. The benchmark score obtained after each iteration is then compared
to the benchmark score obtained without shuffling any variables. The
greater the difference between these two benchmark values, the more
important the variable is for the model \citep{WEI2015399}.

@@

Each drone flight was used to produce a prediction map, as well as a
probability map that indicates the model derived likelihood of the
selected class for every pixel. The low-altitude flight conducted in
Gafanha, Portugal, represents the site with the highest complexity
(Figure~\ref{fig-GafLow}). Among the five vegetation classes on which
the model was trained, four were present on this site, with
Chlorophyceae and Rhodophyceae mixed with the seagrass meadow. There was
also Bacillariophyceae forming biofilms on parts of sediment surface.
Although the seagrass bed was solely composed of \emph{Nanozostera
noltei}, various colors of this specie could be observed from dark green
(corresponding to healthy leaves) to whitish/brown (when leaves were
discolored having lost their pigmentation). Regardless of color,
Magnoliopsida pixels were accurately predicted by the model.

@@

\caption{\label{fig-GafHigh}RGB orthomosaic (Left) and Prediction
(Right) of the high altitude flight of Gafanha, Portugal. The total
extent of this flight was about 1 km² with a resolution of 80 mm per
pixel. Background colors indicate intertidal area (Light Green), land
area (Light Grey) and water (Light Blue). The yellow outline shows the
extent of the low altitude flight of Gafanha presented in
Figure~\ref{fig-GafLow}. The zoom covers an area equivalent to a
10-meter Sentinel-2 pixel size.}

@@

Amoung the high altitude flights, the one acquired over the inner lagoon
of the Ria de Aveiro covered the largest area with approximately 1.5 km²
(Figure~\ref{fig-Boat}). This site was dominated by seagrass and red
macroalgae. The classification provided consistent results, with a
patchy seagrass meadow mixed with red macroalgae on the eastern part of
the site. As shown in the zoom (Figure~\ref{fig-Boat}), the edges of the
meadow visually appears to be colonised by green macroalgae, which the
model agreed with.

@@

The flight over L'Epine in Noirmoutier Island, France
(Figure~\ref{fig-Dike}) was conducted near a dike, which crossed the
northern part of the site from west to east. Alongside this dike, Fucale
brown algae (\emph{Fucus spp.}, \emph{Ascophyllum nodosum}) were
attached to sparse rocks and stranded green algae (\emph{Ulva spp.})
could be observed. Despite the high mixture between Chlorophyceae and
Magnoliopsida these two classes were correctly discriminated by the
classifier.

@@

\caption{\label{fig-Dike}RGB orthomosaic (Top) and Prediction (Bottom)
of L'Epine, France. The total extent of this flight was about 28 000 m²
with a resolution of 80 mm per pixel. Background colors indicate
intertidal area (Light Green) and land area (Light Grey). The zoom
covers an area equivalent to a 10-meter Sentinel-2 pixel size.}

@@

In this study, a key innovation lies in the utilization of drone flights
at two different altitudes (12m and 120m) for constructing the neural
network model. Overall there was a good agreement between the two
altitudes (Figure~\ref{fig-CompareRef}).

@@

There was a slight underestimation for raw reflectance values in the
high-altitude flight, particularly for higher reflectance values
(Figure~\ref{fig-CompareRef} A). Since both flights were conducted over
vegetation areas, the highest reflectance values correspond to the
infrared part of the spectrum. This difference is not present when
reflectance values have been standardized (Equation~\ref{eq-std} ;
Figure~\ref{fig-CompareRef} B).

@@

Model global accuracy was 94.26\% with a Kappa coefficient of 0.92
(Figure~\ref{fig-Validation}).

@@

\caption{\label{fig-Validation}A global confusion matrix on the left is
derived from validation data across each flight, while a mosaic of
confusion matrices from individual flights is presented on the right.
The labels inside the matrices indicate the balanced accuracy for each
class. The labels at the bottom of the global matrix indicate the User's
accuracy for each class, and those on the right indicate the Producer's
Accuracy. The values adjacent to the names of each site represent the
proportion of total pixels from that site contributing to the overall
matrix. Grey lines within the mosaic indicate the absence of validation
data for the class at that site. The table at the bottom summarizes the
Sensitivity, Specificity, and Accuracy for each class and for the
overall model.}

@@

The lowest performing site was Gafanha High (global accuracy of 75.45\%)
whereas Mataducos was the site with the most accurate prediction (global
accuracy of 98.05\%). Overall, the classes Phaeophyceae, Magnoliopsida,
Sediment and Rhodophyceae were correctly classified with a balanced
accuracy of 1, 0.96, 0.96 and 0.91 respectively. Bacillariophyceae was
the least accurate class (accuracy of 0.72 ) mainly due to confusion
with Magnoliopsida and Sediment.

\subsection{Variable importance}\label{variable-importance-1}

@@

\caption{\label{fig-VIP}Variable Importance of the Neural Network
Classifier for each taxonomic class. The longer the slice, the more
important the variable for prediction of each class. The right plot
shows the drone raw and standardised reflectance spectra of each class.
Each slice represents the VIP of both raw and standardised reflectance
combined.}

@@

The spectral bands at 444, 717 and 842 nm of the Micasense camera did
not provide as important information to discriminate any of the
vegetation classes. The band at 531 nm was the only important predictor
for the classifier to accurately predict Chlorophyceae. In fact, at this
wavelength, the Chlorophyceae spectra showed the highest reflectance
among all vegetation classes. The bands at 531 and 740 nm were the most
important predictors for Phaeophyceae, corresponding to the lowest
reflectance among all classes. Bands at 475 and 560 nm were the most
important predictors for Bacillariophyceae and Rhodophyceae,
respectively. Four predictors, ranging from the green (560 nm) to the
RedEdge (705 nm) bands were important to accurately predict
Magnoliopsida.

@@

\caption{\label{fig-upscaling}Kernel density plot showing the proportion
of pixel well classified based on the percent cover of the class in high
altitude flight pixels of Gafanha, Portugal. Each subplot shows all the
pixels of the same classes on the high altitude flight. Percent cover of
classes was retrieved using the result of the classification of the low
altitude flight of Gafanha, Portugal.}

@@

When the percent cover of a given class was 100 \%, coarser high-flight
pixels were well classified for all the classes except for Bare
Sediment, which was only well classified 80\% of the time. This
phenomenon may be attributed to the time gap between the two flights,
allowing for microphytobenthos migration to the surface during low tide,
consequently altering the model's classification from bare sediment to
Bacillariophyceae. A percent cover of at least 80\% was sufficient to
have all the pixels of high altitude flights correctly classified, with
the exception of Magnoliopsida that required a higher percent cover
(\textgreater90 \%) to be well classified. Concerning the probability of
each class, the highest percent cover was needed to confidently predict
Bacillariophyceae. To predict Chlorophyceae with a model likelihood of
0.85, a percent cover of 93 \% was needed, 90 \% for Magnoliopsida, 92
\% for Rhodophyceae and 97 \% for Bacillariophyceae.

@@

The primary objective of this study was to develop a method for the
accurate classification of macrophytes on intertidal mudflats,
specifically focusing on distinguishing between Chlorophyceae (green
macroalgae) and marine Magnoliopsida (seagrasses) using multispectral
drone data. The ability to differentiate between various types of
vegetation plays a critical role in ecological monitoring and coastal
management \citep{WFD2000}. By distinguishing between seagrasses and
macroalgae, our approach facilitates targeted conservation strategies,
enabling more effective preservation and restoration efforts in coastal
ecosystems. The discrimination of seagrasses from green macroalgae
presents two main challenges \citetext{\citealp[
]{oiry2021using}; \citealp[
]{bannari2022}; \citealp{veettil2020opportunities}}. First these two
macrophytes have a similar pigment composition: chlorophyll-a (common to
all vegetation types), chlorophyll-b (an additional photosynthetic
pigment), and accessory carotenoids such as zeaxanthin, lutein and
neoxanthin (Figure~\ref{fig-Pigm}). Second, seagrass and green
macroalgae frequently co-occur in intertidal areas, and can intermingle
within a remote sensing pixel if the spatial resolution is too low.
Here, the issue of intra-pixel mixing was resolved thanks to the very
high spatial resolution of the drone, and the risk of spectral confusion
was avoided thanks to the efficiency of the multispectral classifier.
Our results confirm a recent study based on \emph{in situ} radiometry,
where it was demonstrated that a radiometer with at least eight spectral
bands ranging from 500 to 850 nm and including a green band at 530 nm
and a RedEdge band at 730 nm was crucial to accurately discriminate
green macrophytes \citep{Davies2023}.

@@

Meeting these two criteria, the Micasense RedEdge-MX DUAL camera used in
this study, enabled the classifier to achieve 97\% of accuracy between
these two classes (Figure~\ref{fig-ValidationGreen}). Even if the
pigment composition of green macrophytes is similar, differences in the
spectral shape can be observed, with green algae having a higher
reflectance peak at 560 nm as well as a higher NIR plateau than seagrass
(Figure~\ref{fig-vegetation}). Such differences were previously
attributed to differences in pigments concentration and/or ratios
\citep{bargain2013seasonal}, cellular structure as well as in the
orientation of the plant over the sediment surface \citetext{\citealp[
]{beach1997vivo}; \citealp[
]{kirk1994light}; \citealp{hedley2018influence}}.

The variable importance analysis (Figure~\ref{fig-VIP}) identified that
the band at 531 nm was the most important for accurately identifying
Chlorophyceae. In fact, at this wavelength, Chlorophyceae exhibited the
highest reflectance among all other classes, highlighting the difference
in carotenoids to chlorophyll-a ratios between seagrasses and green
macroalgae \citep{repolho2017seagrass}.

Concerning Phaeophyceae, the thick cell walls of these macroalgae
\citep{charrier2021growth} make it more reflective in the infrared part
of the spectra \citep{Slaton2001} whereas the presence of fucoxanthin
and zeaxanthin result in a low reflectance in the visible region
(Figure~\ref{fig-Pigm} ; Figure~\ref{fig-VIP}). These two key features
have been identified by the Neural Network as the two principal
predictors to accurately identify brown algae (Figure~\ref{fig-VIP}).
Similarly, the presence of phycoerythrin and phycocyanin in Rhodophyceae
contribute to the lowest reflectance among all classes in the spectral
range from 560 to 615 nm (Figure~\ref{fig-VIP}). Indeed the band at 560
nm has been identified as important for identifying this class, likely
due to phycoerythrin absorption at this wavelength. Regarding
Bacillariophyceae, the VIP analysis (Figure~\ref{fig-VIP}) indicated
that 475 nm was the most important predictor for this class. Indeed, the
reflectance at 475 nm was higher for Bacillariophyceae than for any
other vegetation class (Figure~\ref{fig-vegetation}), very likely due to
the low biomass (and associated concentration of blue-absorbing
pigments) of microphytobenthos biofilms compared to seagrass and
macroalgae.

@@

While comparing the reflectance of both altitudes (12 m and 120 m), it
was observed that there is an underestimation of the infrared part of
the spectra in the high-altitude dataset (Figure~\ref{fig-CompareRef}).
Such disparity in infrared reflectance may stem from temporal
differences between the flights, possibly resulting in a slightly drier
intertidal area and consequently higher infrared reflectance. This
disparity poses an issue for the methodology followed in the present
study, relying solely on one flight height for training. To address this
issue, we employed min/max standardized reflectance spectra as
predictors for the model (Equation~\ref{eq-std} ; \citep{Cao2017}). This
approach allowed us to eliminate the slight reflectance difference
between the flights (Figure~\ref{fig-CompareRef} B) and to focus on the
shape of the spectra in the visible part of the electromagnetic spectra,
where different pigmentation are associated to taxonomic dignostic
features. This was a key feature in building a model that could reliably
predict vegetation across geographical sites and seasons. It enabled
consistent prediction of vegetation classes across variations in biomass
and variability in light conditions \citetext{\citealp[
]{fyfe2003spatial}; \citealp[
]{COSTA2021107018}; \citealp{piaser2023impact}}.

We found 90 \% of seagrass cover is necessary for confident prediction
of its presence (Figure~\ref{fig-upscaling}). This highlights a
limitation of the methodology used to construct the training dataset for
the model. The dataset used to train our algorithm was composed
exclusively of pure pixels, which resulted in the model's reduced
confidence when faced with lower percentages of seagrass cover. Also,
intertidal seagrasses exhibit marked phenology, with varying pigment
composition throughout the year \citetext{\citealp[
]{bargain2013seasonal}; \citealp{legare2022remote}}. Since the training
dataset has been made using well developed seagrass meadows, this model
may be less accurate outside of the seasonal seagrass peak of biomass.
Further investigation is required to evaluate the accuracy of the method
along different periods of the year.

@@

Climate change, global warming, eutrophication, alien and invasive
species development, coastal erosion, and sealevel rise are expected to
continue impacting coastal ecosystems in the future
\citetext{\citealp{SCHIBALSKI2022101414}; \citealp[
]{holon2018predictive}; \citealp{marquet2024global}}. and the demand for
meaningful and efficient monitoring of coastal habitats has never been
higher\citetext{\citealp[ ]{muller2018satellite}; \citealp[
]{villalobos2023remote}; \citealp{oiry2021using}}. Our findings,
particularly the improved discrimination of seagrass and green
macroalgae from other intertidal vegetation classes, highlight the
potential of drone-based remote sensing to support diverse applications,
from conservation of biodiversity to climate change adaptation
strategies.

Because of coastal eutrophication, macroalgal blooms are becoming
increasingly common in many regions around the world \citetext{\citealp[
]{sutton2011european}; \citealp{ye2011green}}. These blooms can have
negative impacts on human health and local economic activities,
including human health, fishing and aquaculture, tourism, and
recreational activities \citetext{\citealp[
]{villares1999nitrogen}; \citealp{ye2011green}}. The first green tide
events (i.e.~bloom of green macroalgae of the genus \emph{Ulva}) were
reported in Brittany, France, back in the 1970s and have since been a
concern for local stakeholders and economic activities
\citep{menesguen2018marees}. Some regions of the world have witnessed an
increase in brown macroalgae blooms, predominantly involving algae of
the genus \emph{Sargassum} washing along the Caribbean coastlines
\citep{louime2017sargassum}, and more recently \emph{Rugulopteryx
okamurea} in southern Europe \citep{Roca2022}. Satellite remote sensing
has proven to be a valuable tool for mapping the spatial and temporal
extent of macroalgal blooms worldwide. However, due to limitations in
spatial resolution, it can only effectively map well-developed blooms
\citetext{\citealp[ ]{rs13081408}; \citealp[
]{klemas2012remote}; \citealp{haro2023biointertidal}}. High spatial
resolution drone imagery, coupled with an accurate classification
algorithm, could be used to map the early stages of macroalgal blooms in
areas known to have regular blooms or in new sites. Indeed, this
approach could provide early warning alerts to local managers.
